Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-09 sftrends_manu.Rnw
obsdat fig added

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Three representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

@@

>>>>

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Representative stations from  three locations in the Delta and Suisun Bay were used: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

@@

<<obsdat, fig.cap = 'Observed \\ac{din} from ten stations in the upper \\ac{sfe} Delta.  Data were collected monthly and evaluated with \\ac{wrtds} models from 1976 to 2013. Note different y-axis scales.  See \\cref{fig:delt_map} for station locations.', fig.height = 8, fig.width = 6, cache = T, out.width = '0.8\\textwidth'>>=
data(delt_dat)

missvals <- c('1996-01-01', '2004-04-01') %>% 
  as.Date
missvals <- seq(missvals[1], missvals[2], by = 'month')
missvals <- data.frame(Site_Code = 'D19', Date = missvals, nh = 0, no23 = 0)

toplo <- delt_dat %>% 
  select(Site_Code, Date, nh, no23) %>% 
  rbind(., missvals) %>% 
  gather('var', 'val', nh:no23) %>% 
  mutate(Site_Code = factor(Site_Code, 
    levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3'))
    )

mythm <- theme_minimal() +
  theme(
    legend.position = 'top', 
    legend.title = element_blank(),
    panel.border = element_rect(colour = "black", fill = NA, size = 0.5),
    axis.ticks.x = element_line(),
    axis.ticks.y = element_line(),
    axis.title.x = element_blank(),
    axis.text.y = element_text(size = 6),
    axis.ticks.length = unit(.1, "cm"), 
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank()
  )

cols <- RColorBrewer::brewer.pal(9, 'Blues')[c(4, 7)]
ylab <- expression(paste('DIN (mg ', L^-1, ')'))

leglabs <- list(
  expression(NH [4] ^ '+'),
  expression(paste(NO [2] ^ '-', ' / ', NO [3] ^ '2-'))
  )

ggplot(toplo, aes(x = Date)) +
  geom_area(aes(y = val, group = var, fill = var), stat = 'identity', position = 'stack') + 
  scale_x_date(expand = c(0, 0)) + 
  scale_y_continuous(ylab, expand = c(0, 0)) + 
  facet_grid(Site_Code~., scales = 'free_y') +
  scale_fill_manual(values = cols, labels = leglabs) +
  mythm
